Future challenges for Research Software Engineering

### 2025-09-28 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

Here, we present the results of a workshop discussion (notes are archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the German RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how RSEng should, or could, develop.
Since this paper focuses on the outputs of the deRSE25 conference, we present our thoughts and perspectives in a largely German-focused context.
At the same time, we recognise, the wider international growth of RSEng as a very important development
for people undertaking software development within the research community.
Much of our work and our discussions are also building on wider perspectives for the future of RSEng that are currently under
investigation among RSEng communities in other countries.
While taking a German-focused view in this paper, we therefore aim to highlight resources and materials from
the wider international community where we feel this is relevant.

It is already difficult to structure the spectrum of RSE tasks and responsibilities
today given the spectrum of research software~\cite{Hasselbring2024}.
It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
In order to structure the discussion, we decided to consider
the evolution of the term RSE and its definition,
the growth and development of the RSE community,
the impact of automation and generative artificial intelligence in RSE work,
the impact of complexity and usability in research software,
ethical and social considerations,
the expected transformations in academic institutions,
and the potential impact of alternative career paths for RSEs
and new groups of RTPs (\emph{Research Technical Professional}s).
This work supplements other opinion pieces developed in parallel
in the SE community~\cite{ChueHong2025,Sochat2024Infra,Bencomo2024AEBoK,OliveiraJr2024,Druskat2025,Carleton2022},
which where not known to the workshop participants; remarkably,
many of the projections made by the participants overlap with predictions made
by the SE community in ``Research Software Engineering in 2030''~\cite{Katz2023}.
We conclude the paper with a summary and with our wishes and recommendations.

>>>>

Here, we present the results of a workshop discussion (notes are archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers, but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the German RSE community.

The workshop was held as an open discussion of 45 minutes between the participants.
As can be seen from the workshop material~\cite{goth2025_WSPad}, we prepared some slides to steer the discussion
towards the time frame of 2035.
For a start we prepared questions that we openly asked to the audience,
while also allowing the participants to meander towards questions of their interest.
The prepared questions stem from internal discussions among the workshop hosts.
It started from discussions whether the recently defined competencies of RSEs~\cite{Goth2024}
are just snapshots of our time, or whether they can constitute something that stands the test of time.
Then we asked ourselves how the community of RSEs would interact with some emerging trends.
The collected material was further discussed among the authors in follow-up meetings, structuring and refining the ideas, adding new ideas as described in each of the following sections.

Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how RSEng should, or could, develop.
Since this paper focuses on the outputs of the deRSE25 conference, we present our thoughts and perspectives in a largely German-focused context.
At the same time, we recognise, the wider international growth of RSEng as a very important development
for people undertaking software development within the research community.
Much of our work and our discussions are also building on wider perspectives for the future of RSEng that are currently under
investigation among RSEng communities in other countries.
While taking a German-focused view in this paper, we therefore aim to highlight resources and materials from
the wider international community where we feel this is relevant.

It is already difficult to structure the spectrum of RSE tasks and responsibilities
today, given the spectrum of research software~\cite{Hasselbring2024}.
It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
In order to structure the workshop discussion for this publication, we decided to consider
the evolution of the term RSE and its definition,
the growth and development of the RSE community,
the impact of automation and generative artificial intelligence in RSE work,
the impact of complexity and usability in research software,
ethical and social considerations,
the expected transformations in academic institutions,
and the potential impact of alternative career paths for RSEs
and new groups of RTPs (\emph{Research Technical Professional}s).
This roughly mirrors the questions asked to the audience, where the entire workshop content is condensed into these topics.

This work supplements other opinion pieces developed in parallel
in the SE community~\cite{ChueHong2025,Sochat2024Infra,Bencomo2024AEBoK,OliveiraJr2024,Druskat2025,Carleton2022},
which where not known to the workshop participants; remarkably,
many of the projections made by the participants overlap with predictions made
by the SE community in ``Research Software Engineering in 2030''~\cite{Katz2023}.
We conclude the paper with a summary and with our wishes and recommendations.
